---
title: "PA1_template"
author: "Daniel Cialdella - dcialdella@gmail.com"
date: "August 18, 2016"
output: html_document
---


## Reproducible Research Course 5 - Week 2

```{r setup, include=FALSE, echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```


### Commit containing full submission.</b><br>
1.    Code for reading in the dataset and/or processing the data<br>
2.    Histogram of the total number of steps taken each day<br>
3.    Mean and median number of steps taken each day<br>
4.    Time series plot of the average number of steps taken<br>
5.    The 5-minute interval that, on average, contains the maximum number of steps<br>
6.    Code to describe and show a strategy for imputing missing data<br>
7.    Histogram of the total number of steps taken each day after missing values are imputed<br>
8.    Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends<br>
9.    All of the R code needed to reproduce the results (numbers, plots, etc.) in the report<br>



### Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement -- a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.


### About the data (Input File)

The data for this assignment can be downloaded from the course web site:
  Dataset: Activity monitoring data [52K] - 17568 records / 3 columns
    
The variables included in this dataset are:
 steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
 date: The date on which the measurement was taken in YYYY-MM-DD format
 interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.




### 1. Loading and preprocessing the data

#### 1.1 Load the data (i.e. read.csv())
```{r P1.1, echo=TRUE}
activity  <- read.csv("activity.csv")
```

#### 1.2 Process/transform the data (if necessary) into a format suitable for your analysis


Pre-calculate date with valid data, see kind of data for activity2 (without NA)

```{r P1.2b, echo=TRUE}
dates <- unique( strptime(activity$date, "%Y-%m-%d") )

activity2 <- na.omit(activity)
dates2    <- unique( strptime(activity2$date, "%Y-%m-%d") ) 

min(dates2)
max(dates2)

dim(activity2)
head(activity2,10)
```




### 2. What is mean total number of steps taken per day?

####  2.1 Calculate the total number of steps taken per day

Calculating activities by day.
```{r P2.1, echo=TRUE}
activitybydaySum <- aggregate( list(activity2$steps),  list(activity2$date), sum )
names(activitybydaySum) = c("date","steps")

activitybydayAver  <- aggregate( list(activity2$steps),  list(activity2$date), mean )
names(activitybydayAver) = c("date","steps")

head(activitybydaySum, 20)
summary(activitybydaySum$steps)
```


Graphic showing Activities by day
```{r P2.1b, echo=TRUE}
plot( activitybydaySum$date, activitybydaySum$steps , main="Sum Activity (day by day) ", lwd=1, xlab="Date")
lines( activitybydaySum$date, activitybydaySum$steps , type="l")
```


#### 2.2 If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day

I understand it but I'll do it anyway, showing info in BarPlot and in a histogram (100 parts)
```{r P2.2, echo=TRUE}
hist(activitybydaySum$steps, main="Histogram standard", labels=TRUE, ylim=c(0,30))
```

Slides by 10
```{r P2.2b, echo=TRUE}
hist(activitybydaySum$steps, breaks=10, main="Histogram 10 Slices", labels=TRUE, ylim=c(0,20))
```

Slides by 20
```{r P2.2c, echo=TRUE}
hist(activitybydaySum$steps, breaks=20, main="Histogram 20 Slices", labels=TRUE, ylim=c(0,12))
```



#### 2.3 Calculate and report the mean and median of the total number of steps taken per day

Showing same graph, with MEAN and MEDIAN now.
```{r P2.3, echo=TRUE}
m1 <- mean( activitybydaySum$steps )
m2 <- median( activitybydaySum$steps )

plot( activitybydaySum$date, activitybydaySum$steps , main="Sum Activity by day", lwd=1, xlab="Dates")
lines( activitybydaySum$date, activitybydaySum$steps , type="l")
abline(h=m1, col=2)
```

Printing MEAN and MEDIAN, very very near values... strange
```{r P2.3b, echo=TRUE}
print(m1)
print(m2)
```




### 3. What's the average daily activity pattern?

#### 3.1 Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)


```{r P3.1, echo=TRUE}
AvgActbyInterval <- tapply(activity2$steps,activity2$interval,mean, na.rm=TRUE)

MaxValue <- max( data.frame( AvgActbyInterval ) )

plot(names(AvgActbyInterval), AvgActbyInterval, type="l", main = "Num of Act by Interval", xlab="5-minute Intervals")
abline(h=MaxValue, col=2)
```


#### 3.2 Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

Max Value by Timing and the Interval Associated with it.
```{r P3.2, echo=TRUE}
print (MaxValue)
```



### 4. Imputing missing values

#### 4.1 Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

Review missing values( VALID ? ):
```{r P4.1}
table( ! is.na(activity$steps) )
```
The FALSE records are the invalid ones.


#### 4.2 Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Create a new objects with "MEAN" (activityConNA)
Create a new object with Means (activityPromedio)
Review one/one record (activityConNA) and replace data with Means (activityPromedio).
I have to obtain the number of records 288 (actitivycantidad), needed to calculate the A/B.

```{r P4.2, echo=TRUE}
activityConNA    <- activity
activityPromedio <- tapply(activity$steps, activity$interval,mean, na.rm=TRUE)
actitivycantidad <- nrow( activityPromedio )
```

#### 4.3 Create a new dataset that is equal to the original dataset but with the missing data filled in.


```{r P4.3, echo=TRUE}
for (i in which(is.na(activityConNA)))
    {
    activityConNA[i,1] <- activityPromedio[    ( (i-1) %% actitivycantidad )    +1]
    }
```


#### 4.4 Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r P4.4, echo=TRUE}
hist( tapply(activityConNA$steps, activityConNA$date,sum), main = "Plot 4.4", xlab="Steps", ylim=c(0,40), labels=T )
```

```{r P4.4b, echo=TRUE}
hist( tapply(activityConNA$steps, activityConNA$date,sum), main = "Plot 4.4 (20)", xlab="Steps", ylim=c(0,20), labels=T , breaks=20)
```



### 5. Are there differences in activity patterns between weekdays and weekends?

#### 5.1 Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

Data with filled in will be used for this point. Calculate DiaSemana (Saturday and Sunday), then prepare a new vector for temporary storage and at the end join original data (activityConNA) with temporary vector (DiaSemana2). 

In Spanish I have to use "Sabado and Domingo" terms. I use four terms to confirm that in Spanish or English all will work fine. A new column is created to store the class (normal day or weekend day)
```{r P5.1, echo=TRUE}

# take care about IDIOM. may be in other languaje will fail. and in lowercase.
DiaSemana  <- c("sabado","domingo","saturday","sunday","sabato", "domenica")

# new vector for calculated data
DiaSemana2 <- c()

# will work with spanish weekend names ?
for (i in 1:length(activityConNA$steps))
    {
    if (any(weekdays(as.Date(activityConNA[i,2])) %in% DiaSemana)) 
        {
            ##add "weekend"" to new column for that row
            DiaSemana2[i] <- "weekend"  }
            ##else add "weekday"" to new column for that row
        else { DiaSemana2[i] <- "weekday"
         }
}

# add another column with the kind of day, not the best way... but worked.
activityConNA <- cbind(activityConNA,DiaSemana2)
```

#### 5.2 Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r P5.2, echo=TRUE}

# I use Mxxxx as temporary variables in memory
m5     <- split(activityConNA, activityConNA$DiaSemana2)

# d1 to 5, Lunes, martes, miercoles, jueves, viernes
d1a5 <- do.call(rbind.data.frame,m5[1])
d6y7 <- do.call(rbind.data.frame,m5[2])

# average of data, table apply
avgd1a5 <- tapply(d1a5$steps,d1a5$interval,mean)
avgd6y7 <- tapply(d6y7$steps,d6y7$interval,mean)

# build 2 dataframses to prepare data to be plotted
paramd <- as.data.frame(cbind(avgd1a5, as.numeric(  as.character(  names(avgd1a5) ))))
paramw <- as.data.frame(cbind(avgd6y7, as.numeric(  as.character(  names(avgd6y7) ))))

# actitivycantidad = 288.
dfparamd <- cbind(paramd, rep("weekend", actitivycantidad ))
dfparamw <- cbind(paramw, rep("weekday", actitivycantidad ))

# Column names, 
colnames(dfparamd) <- c("avgSteps", "interval","dayweek")
colnames(dfparamw) <- c("avgSteps", "interval","dayweek")

AllInOnedata <- rbind(dfparamd, dfparamw)

```



##### Final Plots (2 - comparative weekday vs. weekend)

```{r pressure, echo=FALSE}
library(lattice)

xyplot(avgSteps ~ interval | dayweek, AllInOnedata, layout=c(1,2), type="l", main = "Compare WeekDay/WeekEnd", xlab="Steps (/5)", ylab="Average", col=c("green","red") )

```


#### EOF
